---
title: "Data Wrangling"
format:
  revealjs:
    include-in-header:
      - text: |
          <link rel="stylesheet" href="https://cdn.jsdelivr.net/npm/bootstrap@5.3.0/dist/css/bootstrap.min.css">
    theme: [default, clean]
    df-print: paged
    smaller: true
    slide-number: true
    header: 
    header-logo: images/informal_seal_transparent.webp
    self-contained: true
code-annotations: select
slide-level: 3
---

```{css}
        body {
            font-family: 'Segoe UI', Tahoma, Geneva, Verdana, sans-serif;
            margin: 20px;
            background-color: #f5f5f5;
        }
        
        .function-box {
            background: white;
            border: 2px solid #2c5aa0;
            border-radius: 12px;
            max-width: 600px;
            margin: 20px auto;
            box-shadow: 0 4px 12px rgba(0,0,0,0.15);
            overflow: hidden;
        }
        
        .function-header {
            background: linear-gradient(135deg, #2c5aa0, #1e3d6f);
            color: white;
            padding: 15px 20px;
            text-align: center;
        }
        
        .function-name {
            font-size: 24px;
            font-weight: bold;
            font-family: 'Courier New', monospace;
            margin: 0;
        }
        
        .function-syntax {
            font-size: 14px;
            font-family: 'Courier New', monospace;
            margin-top: 8px;
            opacity: 0.9;
        }
        
        .function-body {
            padding: 20px;
        }
        
        .description {
            margin-bottom: clamp(12px, 2.5vw, 20px);
            color: #444;
            line-height: 1.5;
            font-size: clamp(12px, 2.2vw, 16px);
        }
        
        .arguments-title {
            font-size: clamp(14px, 3vw, 20px);
            font-weight: bold;
            margin-bottom: clamp(8px, 2vw, 15px);
            color: #2c5aa0;
            border-bottom: 2px solid #e0e0e0;
            padding-bottom: 5px;
        }
        
        .argument {
            margin-bottom: clamp(8px, 2vw, 15px);
            padding: clamp(8px, 2vw, 15px);
            background: #f8f9fa;
            border-left: 4px solid #2c5aa0;
            border-radius: 0 6px 6px 0;
        }
        
        .arg-name {
            font-family: 'Courier New', monospace;
            font-weight: bold;
            color: #c7254e;
            font-size: clamp(12px, 2.5vw, 18px);
        }
        
        .arg-description {
            margin-top: clamp(3px, 1vw, 8px);
            color: #555;
            line-height: 1.4;
            font-size: clamp(10px, 2vw, 14px);
        }
        
        .arg-type {
            font-style: italic;
            color: #666;
            font-size: clamp(9px, 1.8vw, 13px);
            margin-top: clamp(2px, 0.5vw, 6px);
        }
        
        .example-section {
            margin-top: clamp(15px, 3vw, 25px);
            padding-top: clamp(12px, 2.5vw, 20px);
            border-top: 2px solid #e0e0e0;
        }
        
        .example-title {
            font-size: clamp(12px, 2.5vw, 18px);
            font-weight: bold;
            color: #2c5aa0;
            margin-bottom: clamp(6px, 1.5vw, 12px);
        }
        
        .example-code {
            background: #f4f4f4;
            border: 1px solid #ddd;
            border-radius: 6px;
            padding: clamp(8px, 2vw, 15px);
            font-family: 'Courier New', monospace;
            font-size: clamp(9px, 1.8vw, 14px);
            color: #333;
            white-space: pre-wrap;
            overflow-x: auto;
        }
      .pagedtable table {
        width: auto; /* Adjust table width as needed */
      }
      .pagedtable th, .pagedtable td {
        width: 60px; /* Set width for the first column */
      }
      /* Add more CSS rules for specific columns as needed */
```

### Data wrangling:


-   Split-Apply-Combine operations

-   Long and Wide data

-   Joins

-   Quick-and-dirty tables and plots

-   Recoding


### Tidyverse conventions {.smaller}

We'll make extensive use of `tidyverse` packages for these sections.

Some genre conventions of tidy code

::::: columns
::: {.column width="40%"}
1.  Most functions take a `.data =` argument. Which means so they work well with `|>` and you'll rarely see `$` notation.

2.  A preference for `tidy` data, where each row represents a single case or observation, each column is a variable, and each cell contains a single value.

3.  Some stylistic regularities: like related functions having a `prefix_suffix` naming style. Ex: `str_split`, `str_replace`, and `str_extract`
:::

::: {.column width="60%"}
```{r, eval=FALSE}

library(tidyverse)

url<-'https://www.chesdata.eu/s/CHES_2024_final_v2-c32x.dta'
ches<-haven::read_dta(url)|>
  haven::as_factor()|>
  group_by(country)|>
  
  mutate(seat_share  = seat/sum(seat,na.rm=T),
         vote_share = vote/sum(vote, na.rm=T)
         )|>
  summarise(gallagher = sqrt(.5 * sum((vote_share -seat_share)^2)))|>
  ungroup()|>
  select(country, gallagher)|>
  arrange(-gallagher)|>
  slice_head(n=5)

ches



```
:::
:::::

```{r setup, include=FALSE}

library(tidyverse)
library(tidycensus)

# good default settings for homeworks (suppress warning messages but always show code and output)
knitr::opts_chunk$set(echo = TRUE, 
                      max.print=20,
                      rows.print=10,
                      warning =FALSE,
                      message=FALSE)



```

### Problem {.smaller}

I'm interesting in tracking the annual number of civilians killed by country in the first 6 months of 2025

. . .

I've got a collection of "candidate events" from the UCDP. Here's an example of the formatting:

:::::: columns
:::: {.column width="30%"}
::: incremental
-   My raw data is measured at the event level

-   But I want to aggregate data by country
:::
::::

::: {.column width="70%"}
```{r, echo=FALSE, paged.print=FALSE}


url<-'https://ucdp.uu.se/downloads/candidateged/GEDEvent_v25_01_25_06.csv'

ged<-read_csv(url)

set.seed(400)
ged|>
  filter(deaths_civilians>0)|>
  select(country, conflict_name,date_start, source_headline,  deaths_civilians)|>
  mutate(source_headline = paste(source_headline))|>
  slice_sample(n=5)|>
  knitr::kable(table.attr = 'data-quarto-disable-processing="true"')|>
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)


```
:::
::::::

### Split-Apply-Combine operations

-   Many analyses will require some variation on this basic pattern:

::: incremental
-   **Split** data into meaningful groups

-   **Apply** some statistical calculation for each group (average, range, variance, etc.)

-   **Combine** the results into a new summary table

-   while these sorts of operations could be performed in a loop, or using base functions like `aggregate`, we'll use the `dplyr` package here
:::

### Split-Apply-Combine

```{dot}
//| fig-cap: "Example of split-apply-combine process for a group-wise average"

digraph SplitApplyCombine {
    rankdir=LR;
    node [shape=box];
    "Original Data" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR>
                <TD>Group</TD>
                <TD>Value</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">3</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">3</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">5</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">7</TD>
            </TR>
           <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">2</TD>
            </TR>
            <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">1</TD>
            </TR>
            <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">1</TD>
            </TR>
            <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">9</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700">C</TD>
                <TD bgcolor="#FFD700" >2</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700" >C</TD>
                <TD bgcolor="#FFD700">1</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700">C</TD>
                <TD bgcolor="#FFD700">1</TD>
            </TR>
            <TR><TD COLSPAN="2">...</TD></TR>

        </TABLE>
    >]
    
    "Final Result" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR>
                <TD >Group</TD>
                <TD >Value</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">4.5</TD>
            </TR>
            <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">3.25</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700">C</TD>
                <TD bgcolor="#FFD700">1.33</TD>
            </TR>
        </TABLE>
    >]
    
    "Group A"[shape=box, style=filled, fillcolor=white];
    "Group B"[shape=box, style=filled, fillcolor="#89F336"];
    "Group C"[shape=box, style=filled, fillcolor="#FFD700"];
    "Result A"[shape=box, style=filled, fillcolor=white];
    "Result B"[shape=box, style=filled, fillcolor="#89F336"];
    "Result C"[shape=box, style=filled, fillcolor="#FFD700"];
    

    // Initial data node
    //"Original Data" [shape=cylinder, label="Original Data\n(e.g., DataFrame)"];

    // The Split phase represented by a subgraph
    subgraph cluster_split {
        label="1. Split";
        style=filled;
        color=lightgrey;
        node [shape=box, style=filled, fillcolor=white];
        
        "Original Data" -> "Group A";
        "Original Data" -> "Group B";
        "Original Data" -> "Group C";
    }

    // The Apply phase represented by a subgraph
    subgraph cluster_apply {
        label="2. Apply";
        style=filled;
        color=lightgrey;
        node [shape=note, style=filled, fillcolor=white];
        
        "Group A" -> "Result A" [label="  mean(Value)"];
        "Group B" -> "Result B" [label="  mean(Value)"];
        "Group C" -> "Result C" [label="  mean(Value)"];
    }

    // The Combine phase represented by a subgraph
    subgraph cluster_combine {
        label="3. Combine";
        style=filled;
        color=lightgrey;
        node [shape=cylinder, style=filled, fillcolor=white];
        
        "Result A" -> "Final Result";
        "Result B" -> "Final Result";
        "Result C" -> "Final Result";
    }
}

```

### dplyr: group_by

::::: columns
::: {.column width="50%"}

`group_by(.data, ...)`

:   Turns `.data` into a grouped table where each group is defined by the variables listed in `...`

-   This doesn't do much on its own, but subsequent dplyr functions will be performed on the groups you've created. You can think of this as the "split" step.
:::

::: {.column width="50%"}
```{r, paged.print=FALSE}
library(dplyr)
df<-data.frame(group = c("A", "A", "A", "A", 
                         "B", "B",  "B", "B",
                         "C", "C", "C"),
               value = c(3, 3, 5, 7, 2, 1, 1,9, 2, 1, 1)
               )

df_grouped<-df|>
  group_by(group)


# data looks the same after this step:
df_grouped

```
:::
:::::

## dplyr summarize

::::: columns
::: {.column width="50%"}

`summarize(.data, ...)`

:   applies one or more functions defined by `...` to each group in `.data` and then combines the results in to a single data frame.

-   You can think of this as performing both the "apply" and "combine" steps.
:::

::: {.column width="50%"}
```{r}

df_grouped<-df|>
  group_by(group)|>
  summarize(mean_value = mean(value),
            sd_value = sd(value)
            
            )

df_grouped


```
:::
:::::

## dplyr group_by + summarize

```{dot}
//| fig-cap: "Example of group_by and summarize"

digraph SplitApplyCombine {
    rankdir=LR;
    node [shape=box];
    "Original Data" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR>
                <TD>group</TD>
                <TD>value</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">3</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">3</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">5</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">7</TD>
            </TR>
           <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">2</TD>
            </TR>
            <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">1</TD>
            </TR>
            <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">1</TD>
            </TR>
            <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">9</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700">C</TD>
                <TD bgcolor="#FFD700" >2</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700" >C</TD>
                <TD bgcolor="#FFD700">1</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700">C</TD>
                <TD bgcolor="#FFD700">1</TD>
            </TR>
            <TR><TD COLSPAN="2">...</TD></TR>

        </TABLE>
    >]
    
    "Final Result" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR>
                <TD >group</TD>
                <TD >mean_value</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">4.5</TD>
            </TR>
            <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">3.25</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700">C</TD>
                <TD bgcolor="#FFD700">1.33</TD>
            </TR>
        </TABLE>
    >]
    
    "Group A"[shape=box, style=filled, fillcolor=white];
    "Group B"[shape=box, style=filled, fillcolor="#89F336"];
    "Group C"[shape=box, style=filled, fillcolor="#FFD700"];
    "Result A"[shape=box, style=filled, fillcolor=white];
    "Result B"[shape=box, style=filled, fillcolor="#89F336"];
    "Result C"[shape=box, style=filled, fillcolor="#FFD700"];
    

    // Initial data node
    //"Original Data" [shape=cylinder, label="Original Data\n(e.g., DataFrame)"];

    // The Split phase represented by a subgraph
    subgraph cluster_split {
        label="group_by(group)";
        style=filled;
        color=lightgrey;
        node [shape=box, style=filled, fillcolor=white];
        
        "Original Data" -> "Group A";
        "Original Data" -> "Group B";
        "Original Data" -> "Group C";
    }

    // The Apply phase represented by a subgraph
    subgraph cluster_apply {
        label="summarize(mean_value = mean(value))";
        style=filled;
        color=lightgrey;
        node [shape=note, style=filled, fillcolor=white];
        
        "Group A" -> "Result A" [label="  "];
        "Group B" -> "Result B" [label=" "];
        "Group C" -> "Result C" [label="  "];
            // The Combine phase represented by a subgraph
    subgraph cluster_combine {
        label="";
        style=filled;
        color=lightgrey;
        node [shape=cylinder, style=filled, fillcolor=white];
        
        "Result A" -> "Final Result";
        "Result B" -> "Final Result";
        "Result C" -> "Final Result";
    }
    }


}

```

## dplyr group_by + mutate

An alternative to `summarize` is `mutate`. Mutate works almost the same way, but the final result is then joined to the original data.

```{dot}

digraph SplitApplyCombine {
    rankdir=LR;
    node [shape=box];
    "Original Data" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR>
                <TD>group</TD>
                <TD>value</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">3</TD>
                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">3</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">5</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">7</TD>
            </TR>
           <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">2</TD>
            </TR>
            <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">1</TD>
            </TR>
            <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">1</TD>
            </TR>
            <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">9</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700">C</TD>
                <TD bgcolor="#FFD700" >2</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700" >C</TD>
                <TD bgcolor="#FFD700">1</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700">C</TD>
                <TD bgcolor="#FFD700">1</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700">C</TD>
                <TD bgcolor="#FFD700">5</TD>
            </TR>
            <TR><TD COLSPAN="2">...</TD></TR>

        </TABLE>
    >]
    
    "Mutated Data" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR>
                <TD>group</TD>
                <TD>value</TD>
                <TD>mean_value</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">3</TD>
                <TD bgcolor="#FFFFFF">4.5</TD>
                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">3</TD>
                <TD bgcolor="#FFFFFF">4.5</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">5</TD>
                <TD bgcolor="#FFFFFF">4.5</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">7</TD>
                <TD bgcolor="#FFFFFF">4.5</TD>
            </TR>
           <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">2</TD>
                <TD bgcolor="#89F336">3.25</TD>

            </TR>
            <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">1</TD>
                <TD bgcolor="#89F336">3.25</TD>

            </TR>
            <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">1</TD>
                <TD bgcolor="#89F336">3.25</TD>

            </TR>
            <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">9</TD>
                <TD bgcolor="#89F336">3.25</TD>

            </TR>
            <TR>
                <TD bgcolor="#FFD700">C</TD>
                <TD bgcolor="#FFD700" >2</TD>
                <TD bgcolor="#FFD700">2.25</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700" >C</TD>
                <TD bgcolor="#FFD700">1</TD>
                <TD bgcolor="#FFD700">2.25</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700">C</TD>
                <TD bgcolor="#FFD700">1</TD>
                <TD bgcolor="#FFD700">2.25</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700">C</TD>
                <TD bgcolor="#FFD700">5</TD>
                <TD bgcolor="#FFD700">2.25</TD>
            </TR>
            <TR><TD COLSPAN="3">...</TD></TR>

        </TABLE>
    >]
    
    "Final Result" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR>
                <TD >group</TD>
                <TD >mean_value</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">A</TD>
                <TD bgcolor="#FFFFFF">4.5</TD>
            </TR>
            <TR>
                <TD bgcolor="#89F336">B</TD>
                <TD bgcolor="#89F336">3.25</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFD700">C</TD>
                <TD bgcolor="#FFD700">2.25</TD>
            </TR>
        </TABLE>
    >]
    
    "Group A"[shape=box, style=filled, fillcolor=white];
    "Group B"[shape=box, style=filled, fillcolor="#89F336"];
    "Group C"[shape=box, style=filled, fillcolor="#FFD700"];
    "Result A"[shape=box, style=filled, fillcolor=white];
    "Result B"[shape=box, style=filled, fillcolor="#89F336"];
    "Result C"[shape=box, style=filled, fillcolor="#FFD700"];
    

    // Initial data node
    //"Original Data" [shape=cylinder, label="Original Data\n(e.g., DataFrame)"];

    // The Split phase represented by a subgraph
    subgraph cluster_split {
        label="groupby(group)";
        style=filled;
        color=lightgrey;
        node [shape=box, style=filled, fillcolor=white];
        
        "Original Data" -> "Group A";
        "Original Data" -> "Group B";
        "Original Data" -> "Group C";
    }

    // The Apply phase represented by a subgraph
    subgraph cluster_apply {
        label="mutate(mean_value = mean(value))";
        style=filled;
        color=lightgrey;
        node [shape=note, style=filled, fillcolor=white];
        
        "Group A" -> "Result A" [label="mean"];
        "Group B" -> "Result B" [label="mean"];
        "Group C" -> "Result C" [label="mean"];
            // The Combine phase represented by a subgraph
    subgraph cluster_combine {
        label="";
        style=filled;
        color=lightgrey;
        node [shape=cylinder, style=filled, fillcolor=white];
        
        "Result A" -> "Final Result";
        "Result B" -> "Final Result";
        "Result C" -> "Final Result";
    }
    subgraph cluster_combine {
        label="";
        style=filled;
        color=lightgrey;
        node [shape=cylinder, style=filled, fillcolor=white];
        
        "Final Result" -> "Mutated Data"
        "Original Data" -> "Mutated Data" [label = "Joining group summaries with original data"]
    }
    }


}


```

## dplyr mutate vs. summarize

Mutate will be most useful when we need to do something like mean-centering or calculating a running-sum for each group in a data set. (We'll also sometimes use `mutate` without `group_by` to just add a new column to a dataframe)

::::: columns
::: {.column width="50%"}
```{r}
df_summarized<-df|>
  group_by(group)|>
  summarize(mean_value = mean(value),
            sd_value = sd(value)
            
            )

df_summarized

```
:::

::: {.column width="50%"}
```{r}

df_mutated<-df|>
  group_by(group)|>
  mutate(mean_value = mean(value),
            sd_value = sd(value)
            
            )|>
  # good rule to always run `ungroup` after mutate:

  ungroup()

df_mutated



```
:::
:::::

## Back to the problem

::: incremental
-   Back to the original problem: how do we turn event-level data into data with a total number of civilian deaths for each country?
:::

. . .

::::: columns
::: {.column width="50%"}
-   **Split** by country

-   **Apply** a sum the civilian death toll column in each group

-   **combine** the results into a single data frame
:::

::: {.column width="50%"}
```{r}

url<-'https://ucdp.uu.se/downloads/candidateged/GEDEvent_v25_0_7.csv'

ged<-read_csv(url)

```

```{r, eval=FALSE}
#| code-fold: true
#| code-summary: "Show the code"

ged|>
  group_by(country)|>
  summarise(civilian_deaths = sum(deaths_civilians))

```
:::
:::::

## Some other common dplyr operations

These functions also respect the grouping context, so be careful about how you use them after calling `group_by` because they can give very different results for a grouped data frame:

-   `select`: selects certain columns from a tbl (with a more flexible syntax compared to base R)

-   `filter`: removes rows based on a logical condition

-   `slice_head`/`slice_tail/slice_sample`: take the top rows, bottom rows, or a random sample of rows.

-   `arrange`: sort by one or more values

-   `count`: count the number of rows in the grouping context.

## Some other common dplyr operations

Here's an example of using some other `dplyr` functions:

. . .

::::: columns
::: {.column width="40%"}
-   line 4. remove cases with 0 deaths

-   line 5-6. get the worst single incident for each conflict

-   line 7. select specific some specific columns

-   line 8. sort by civilian deaths in decreasing order

-   line 9. ungroup the result
:::

::: {.column width="60%"}
```{r, paged.print=FALSE}
#| code-line-numbers: true
#| df-print: kable


url<-'https://ucdp.uu.se/downloads/candidateged/GEDEvent_v25_0_7.csv'
ged<-read_csv(url)
ged|>
  filter(deaths_civilians>0)|>
  group_by(conflict_name)|>
  filter(deaths_civilians == max(deaths_civilians))|>
  select(date_start, conflict_name, deaths_civilians)|>
  arrange(-deaths_civilians)|>
  ungroup()



```
:::
:::::

## Problem

The UCDP distinguishes between state based, non-state, and one-sided forms of violence. I'm interested in tracking total deaths by country and violence type.

. . .

We can think of a single "record" of data as looking something like this:

```{dot}

digraph G {

          node [shape=box];
     "Record 1" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ='2'>Single record: one for each country/violence type</TD></TR>

            
            <TR><TD> Country </TD><TD>Ex: Burkina Faso, Mexico, Ukraine, ... </TD></TR>
            <TR><TD> Conflict Type </TD><TD>Ex: non-state, state, one-sided</TD></TR>
            <TR><TD> Est. Deaths </TD><TD>Ex: 296, 376, 6870 </TD></TR>

        
            </TABLE>
            >
            
            ]
}

```

. . .

How should I arrange my data for analysis?

## Complex data and long vs. wide format

This is another common complication: how do we represent complex data 2-dimensional matrix?

A classic example would be panel data, where multiple cases are tracked across multiple years:

-   Country-year data on economic growth

-   Tracking student GPAs across terms

-   Public opinion surveys for the same set of respondents during an entire election

. . .

## Long Data

Putting the records in **long format** would mean putting each case in a row and using repetition to identify related cases:

```{dot}

digraph G {
    rankdir=TB;

    node [shape=box];

          node [shape=box];
     "Record 1" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ='2'>Single record: one for each country/violence type</TD></TR>

            
            <TR><TD> Country </TD><TD>Ex: Burkina Faso, Mexico, Ukraine, ... </TD></TR>
            <TR><TD> Conflict Type </TD><TD>Ex: state-based, non-state, one-sided</TD></TR>
            <TR><TD> Est. Deaths </TD><TD>Ex: 296, 376, 6870 </TD></TR>

        
            </TABLE>
            >
            
            ]
   "Long Format" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4"> Long Format </TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>Conflict Type</TD>
                <TD>Est. Deaths</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Burkina Faso</TD>
                <TD bgcolor="#FFFFFF">state-based</TD>
                <TD bgcolor="#FFFFFF">296</TD>
                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Burkina Faso</TD>
                <TD bgcolor="#FFFFFF">non-state</TD>
                <TD bgcolor="#FFFFFF">6</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Burkina Faso</TD>
                <TD bgcolor="#FFFFFF">one-sided</TD>
                <TD bgcolor="#FFFFFF">58</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Mexico</TD>
                <TD bgcolor="#FFFFFF">state-based</TD>
                <TD bgcolor="#FFFFFF">8</TD>
            </TR>
           <TR>
                <TD bgcolor="#FFFFFF">Mexico</TD>
                <TD bgcolor="#FFFFFF">non-state</TD>
                <TD bgcolor="#FFFFFF">376</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Mexico</TD>
                <TD bgcolor="#FFFFFF">one-sided</TD>
                <TD bgcolor="#FFFFFF">22</TD>
            </TR>
            <TR><TD colspan ="4"> ...more rows</TD></TR>


        </TABLE>
    >]
    

        subgraph cluster_split {
        label="";
        style=filled;
        color=white;
        node [shape=box, style=filled, fillcolor=white];

       "Record 1" ->"Long Format"


     }


   
}


```

## Wide Format

Putting records in *wide format* means using additional columns for each group (either one column per country or per violence type, in the current example)

```{dot}


digraph G {
    rankdir=TB;

    node [shape=box];

          node [shape=box];
     "Record 1" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ='2'>Single record: one for each country/violence type</TD></TR>

            
            <TR><TD> Country </TD><TD>Ex: Burkina Faso, Mexico, Ukraine, ... </TD></TR>
            <TR><TD> Conflict Type </TD><TD>Ex: state-based, non-state, one-sided</TD></TR>
            <TR><TD> Est. Deaths </TD><TD>Ex: 296, 376, 6870 </TD></TR>

        
            </TABLE>
            >
            
            ]
   "Long Format" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4"> Long Format </TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>Conflict Type</TD>
                <TD>Est. Deaths</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Burkina Faso</TD>
                <TD bgcolor="#FFFFFF">state-based</TD>
                <TD bgcolor="#FFFFFF">296</TD>
                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Burkina Faso</TD>
                <TD bgcolor="#FFFFFF">non-state</TD>
                <TD bgcolor="#FFFFFF">6</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Burkina Faso</TD>
                <TD bgcolor="#FFFFFF">one-sided</TD>
                <TD bgcolor="#FFFFFF">58</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Mexico</TD>
                <TD bgcolor="#FFFFFF">state-based</TD>
                <TD bgcolor="#FFFFFF">8</TD>
            </TR>
           <TR>
                <TD bgcolor="#FFFFFF">Mexico</TD>
                <TD bgcolor="#FFFFFF">non-state</TD>
                <TD bgcolor="#FFFFFF">376</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Mexico</TD>
                <TD bgcolor="#FFFFFF">one-sided</TD>
                <TD bgcolor="#FFFFFF">22</TD>
            </TR>
            <TR><TD colspan ="4"> ...more rows</TD></TR>


        </TABLE>
    >]
    
    
       "Wide Format" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4"> Wide Format </TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>state-based</TD>
                <TD>non-state</TD>
                <TD>one-sided</TD>

            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Burkina Faso</TD>

                <TD bgcolor="#FFFFFF">296</TD>
                <TD bgcolor="#FFFFFF">6</TD>
                <TD bgcolor="#FFFFFF">58</TD>

                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Mexico</TD>
                <TD bgcolor="#FFFFFF">8</TD>
                <TD bgcolor="#FFFFFF">376</TD>
                <TD bgcolor="#FFFFFF">22</TD>



            </TR>

            <TR><TD colspan ="4"> ...more rows</TD></TR>

        </TABLE>
    >]

        subgraph cluster_split {
        label="";
        style=filled;
        color=white;
        node [shape=box, style=filled, fillcolor=white];

       "Record 1" ->"Long Format"


     }

       "Record 1" ->"Wide Format"

   
}

```

## Long vs. Wide

:::::: columns
:::: {.column width="50%"}
::: incremental
-   Wide format data is generally easier on the eyeballs, especially when you want people to make comparisons across groups/periods.
-   It's much easier to calculate totals by row or column.
-   But it's less flexible: what if I want to track civilian and non-civilian deaths separately? I need a whole new table.
:::
::::

::: {.column width="50%"}
```{dot}
//| fig-width: 5

digraph G {
    rankdir=TB;

    node [shape=box];

"Wide Format" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4"> Wide Format </TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>state-based</TD>
                <TD>non-state</TD>
                <TD>one-sided</TD>

            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Burkina Faso</TD>

                <TD bgcolor="#FFFFFF">296</TD>
                <TD bgcolor="#FFFFFF">6</TD>
                <TD bgcolor="#FFFFFF">58</TD>

                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Mexico</TD>
                <TD bgcolor="#FFFFFF">8</TD>
                <TD bgcolor="#FFFFFF">376</TD>
                <TD bgcolor="#FFFFFF">22</TD>



            </TR>

            <TR><TD colspan ="4"> ...more rows</TD></TR>

        </TABLE>
    >]

}


```
:::
::::::

## Long vs. Wide

:::::: columns
:::: {.column width="50%"}
::: incremental
-   Long format may be inefficient for storage and transfer because information is repeated.

-   But it's far more flexible: adding combatant death tolls is as simple as adding a new column.

-   Each row is a discrete observation as opposed to a collection of observations.

-   In general, R functions (especially code related to the tidyverse) will work better with long data.
:::
::::

::: {.column width="50%"}
```{dot}
//| fig-width: 5

digraph G {
    rankdir=TB;

    node [shape=box];

 "Long Format" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4"> Long Format </TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>Conflict Type</TD>
                <TD>Est. Deaths</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Burkina Faso</TD>
                <TD bgcolor="#FFFFFF">state-based</TD>
                <TD bgcolor="#FFFFFF">296</TD>
                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Burkina Faso</TD>
                <TD bgcolor="#FFFFFF">non-state</TD>
                <TD bgcolor="#FFFFFF">6</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Burkina Faso</TD>
                <TD bgcolor="#FFFFFF">one-sided</TD>
                <TD bgcolor="#FFFFFF">58</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Mexico</TD>
                <TD bgcolor="#FFFFFF">state-based</TD>
                <TD bgcolor="#FFFFFF">8</TD>
            </TR>
           <TR>
                <TD bgcolor="#FFFFFF">Mexico</TD>
                <TD bgcolor="#FFFFFF">non-state</TD>
                <TD bgcolor="#FFFFFF">376</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Mexico</TD>
                <TD bgcolor="#FFFFFF">one-sided</TD>
                <TD bgcolor="#FFFFFF">22</TD>
            </TR>
            <TR><TD colspan ="4"> ...more rows</TD></TR>


        </TABLE>
    >]
    
    
}


```
:::
::::::

## Long to wide in dplyr

::::::: columns
::: {.column width="50%"}

`pivot_wider(.data, names_from, values_from)`

:   Makes data wider. `names_from` will identify the new column names. `values_from` will identify the values in each cell.
:::

::::: {.column width="50%"}
::: fragment
```{r, paged.print=FALSE}

df<-data.frame(group = c("A", "A", "A",
                         "B", "B",  "B"),
               period = c("First", "Second", "Third", 
                          "First", "Second", "Third"),
               value = c(3, 3, 5, 7, 2, 1)
               )
df


```
:::

::: fragment
```{r, paged.print=FALSE}

df|>
  pivot_wider(names_from = group, values_from = value)

```
:::
:::::
:::::::

## Wide-to-long in dplyr

::::::: columns
::: {.column width="50%"}

`pivot_longer(.data, cols)`

:   Makes data longer. Data in `cols` will form new rows.
:::

::::: {.column width="50%"}
::: fragment
```{r, paged.print=FALSE}

df<-data.frame(period = c("First", "Second", "Third"),
               `A` = c(3, 3, 5),
               `B` = c(7, 2, 1)
               )
df


```
:::

::: fragment
```{r, paged.print=FALSE}

df|>
  pivot_longer(cols=c(A,B))

```
:::
:::::
:::::::

## Back to the problem

How should I arrange my data for analysis?

```{r, paged.print=FALSE}

url<-'https://ucdp.uu.se/downloads/candidateged/GEDEvent_v25_0_7.csv'
ged<-read_csv(url)
ged_by_type=ged|>
  mutate(
    type_of_violence = factor(
    type_of_violence,
    levels = c(1, 2, 3),
    labels = c("State-based", "Non-state", "One-sided")
  ))|>
  group_by(country, type_of_violence)|>
  summarise(est_deaths  = sum(best))

ged_by_type




```

## Back to the problem

Try pivoting this to wide format and then back again.

::::: columns
::: {.column width="50%"}
Pivoting the data from long to wide with one column for each conflict type.

```{r, paged.print=FALSE}
#| code-fold: true
#| code-summary: "Show the code"

ged_wide<-ged_by_type|>
  pivot_wider(names_from = type_of_violence, 
              values_from = est_deaths)
  
ged_wide


```
:::

::: {.column width="50%"}
Putting the wide data back in its original format:

```{r, paged.print=FALSE}
#| code-fold: true
#| code-summary: "Show the code"

ged_long <- ged_wide|>
  pivot_longer(cols = c(`State-based`, `Non-state`, `One-sided`),
               names_to ='violence_type', #  name the new columns
               values_to ='est_deaths'
               )
  
  
ged_long

```
:::
:::::

## Long vs. Wide

-   A general rule: prefer long data while you're processing stuff. You can use wide formats for your "finished products" (the stuff that someone else might need to look at)

-   Keep in mind: Wide and long aren't necessarily discrete categories. There are long-er and wide-er versions of most real-world data, so it's really more of a relative concept.

### Problem

I want to explore the relationship between "rugged terrain" and civil wars since 1945, but doing so requires me to combine data from two different data sets.

:::::: columns
::: {.column width="50%"}
My first source is a list of all countries that had a civil war since 1989.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# downloading battle deaths data
if(!file.exists('ucdp_datasets/BattleDeaths_v25_1.csv')){
  download.file('https://ucdp.uu.se/downloads/brd/ucdp-brd-dyadic-251-csv.zip',
                dest = 'battle_deaths.zip'
                )
  
    unzip('battle_deaths.zip',
        exdir = 'ucdp_datasets')
}


# getting data on civil wars
civil_wars<-read_csv('ucdp_datasets/BattleDeaths_v25_1.csv')|>
  filter(type_of_conflict=='3')|>
  group_by(gwno_a)|>
  summarise(deaths = sum(bd_best))|>
  filter(deaths>1000)|>
  ungroup()|>
  mutate(gwno_a = as.numeric(gwno_a))|>
  mutate(country = countrycode::countrycode(gwno_a, origin='gwn', 
                                            dest='country.name'))|>
  relocate(country)|>
  arrange(country)


civil_wars

```
:::

:::: {.column width="50%"}
::: fragment
My second data source measures the logged % of mountainous terrain in a country:

```{r}
#| code-fold: true
#| code-summary: "Show the code"

rugged<-peacesciencer::rugged|>
  filter(!is.na(gwcode))|>
  mutate(country = countrycode::countrycode(gwcode, origin='gwn', 
                                            dest='country.name'),
         country = ifelse(gwcode == 816, "Vietnam", country))|>
  relocate(country)|>
  select(-ccode, -rugged)|>
  arrange(country)|> 
  rename(rough = newlmtnest)


rugged

```
:::
::::
::::::

## Joins

This will require some kind of join, where two tables are combined based on a matching value in each called the **key.** But there's more than one way to do this, especially when the tables don't have perfectly matching keys.

```{dot}
digraph G {
    rankdir=LR;

    node [shape=box];


   "civil_wars" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4">Civil Wars</TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>Country ID</TD>
                <TD>Est. Deaths</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Afghanistan</TD>
                <TD bgcolor="#FFFFFF">700</TD>
                <TD bgcolor="#FFFFFF">54753</TD>
                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Algeria</TD>
                <TD bgcolor="#FFFFFF">615</TD>
                <TD bgcolor="#FFFFFF">18488</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Angola</TD>
                <TD bgcolor="#FFFFFF">540</TD>
                <TD bgcolor="#FFFFFF">23088</TD>
            </TR>
            <TR><TD colspan ="4"> ...more rows</TD></TR>


        </TABLE>
    >]
    
    
       "rugged" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4">Terrain</TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>Country ID</TD>
                <TD>Log % Rough Terrain</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Afghanistan</TD>
                <TD bgcolor="#FFFFFF">700</TD>
                <TD bgcolor="#FFFFFF">4.2</TD>
                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Albania</TD>
                <TD bgcolor="#FFFFFF">339</TD>
                <TD bgcolor="#FFFFFF">2.3</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Algeria</TD>
                <TD bgcolor="#FFFFFF">615</TD>
                <TD bgcolor="#FFFFFF">2.8</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Angola</TD>
                <TD bgcolor="#FFFFFF">540</TD>
                <TD bgcolor="#FFFFFF">2.4</TD>
            </TR>
            <TR><TD colspan ="4"> ...more rows</TD></TR>


        </TABLE>
    >]
    
    
        "Combined Data" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">


        </TABLE>
    >]

        subgraph cluster_split {
        label="";
        style=filled;
        color=white;
        node [shape=box, style=filled, fillcolor=white];



     }



    
    rugged -> "Combined Data"
    
    civil_wars -> "Combined Data"

   
}

```

## Join types

-   Inner joins preserve only keys that are present in **both** data sets. (Rows with multiple matches are duplicated)

```{dot}

digraph G {
    rankdir=LR;

    node [shape=box];


   "civil_wars" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4">Civil Wars</TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>Country ID</TD>
                <TD>Est. Deaths</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Afghanistan</TD>
                <TD bgcolor="#FFFFFF">700</TD>
                <TD bgcolor="#FFFFFF">54753</TD>
                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Algeria</TD>
                <TD bgcolor="#FFFFFF">615</TD>
                <TD bgcolor="#FFFFFF">18488</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Angola</TD>
                <TD bgcolor="#FFFFFF">540</TD>
                <TD bgcolor="#FFFFFF">23088</TD>
            </TR>
            <TR><TD colspan ="4"> ...more rows</TD></TR>


        </TABLE>
    >]
    
    
       "rugged" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4">Terrain</TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>Country ID</TD>
                <TD>Log % Rough Terrain</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Afghanistan</TD>
                <TD bgcolor="#FFFFFF">700</TD>
                <TD bgcolor="#FFFFFF">4.2</TD>

                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Albania</TD>
                <TD bgcolor="#FFFFFF">339</TD>
                <TD bgcolor="#FFFFFF">2.3</TD>

            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Algeria</TD>
                <TD bgcolor="#FFFFFF">615</TD>
                <TD bgcolor="#FFFFFF">2.8</TD>

            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Angola</TD>
                <TD bgcolor="#FFFFFF">540</TD>
                <TD bgcolor="#FFFFFF">2.4</TD>

            </TR>
            <TR><TD colspan ="4"> ...more rows</TD></TR>


        </TABLE>
    >]
    
    
        "joined" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4">Combined</TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>Country ID</TD>
                <TD>Log % Rough Terrain</TD>
                <TD>Est. Deaths</TD>

            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Afghanistan</TD>
                <TD bgcolor="#FFFFFF">700</TD>
                <TD bgcolor="#FFFFFF">4.2</TD>
                <TD bgcolor="#FFFFFF">54753</TD>

                
            </TR>
            <TR>
              <TD colspan ="4">Albania dropped!</TD>


            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Algeria</TD>
                <TD bgcolor="#FFFFFF">615</TD>
                <TD bgcolor="#FFFFFF">2.8</TD>
                <TD bgcolor="#FFFFFF">18488</TD>

            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Angola</TD>
                <TD bgcolor="#FFFFFF">540</TD>
                <TD bgcolor="#FFFFFF">2.4</TD>
                <TD bgcolor="#FFFFFF">23088</TD>

            </TR>
            <TR><TD colspan ="4"> ...more rows</TD></TR>


        </TABLE>
    >]

        subgraph cluster_split {
        label="";
        style=filled;
        color=white;
        node [shape=box, style=filled, fillcolor=white];



     }



    
    rugged -> joined
    
    civil_wars -> joined

   
}

```



## Left/Right join

-   A `left_join` will take all data from the left hand side and all matching data on the left. A `right_join` will do the reverse. ("left" vs. "right" is entirely a matter of which table you reference first in the join command)


```{dot}

digraph G {
    rankdir=TD;

    node [shape=box];

 "rugged" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4">Terrain (RHS)</TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>Country ID</TD>
                <TD>Log % Rough Terrain</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Afghanistan</TD>
                <TD bgcolor="#FFFFFF">700</TD>
                <TD bgcolor="#FFFFFF">4.2</TD>

                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Albania</TD>
                <TD bgcolor="#FFFFFF">339</TD>
                <TD bgcolor="#FFFFFF">2.3</TD>

            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Algeria</TD>
                <TD bgcolor="#FFFFFF">615</TD>
                <TD bgcolor="#FFFFFF">2.8</TD>

            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Angola</TD>
                <TD bgcolor="#FFFFFF">540</TD>
                <TD bgcolor="#FFFFFF">2.4</TD>

            </TR>
            <TR><TD colspan ="4"> ...more rows</TD></TR>


        </TABLE>
    >]

   "civil_wars" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4">Civil Wars (LHS)</TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>Country ID</TD>
                <TD>Est. Deaths</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Afghanistan</TD>
                <TD bgcolor="#FFFFFF">700</TD>
                <TD bgcolor="#FFFFFF">54753</TD>
                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Algeria</TD>
                <TD bgcolor="#FFFFFF">615</TD>
                <TD bgcolor="#FFFFFF">18488</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Angola</TD>
                <TD bgcolor="#FFFFFF">540</TD>
                <TD bgcolor="#FFFFFF">23088</TD>
            </TR>
            <TR><TD colspan ="4"> ...more rows</TD></TR>


        </TABLE>
    >]
    
    
      
    
    
        "joined" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4">Terrain</TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>Country ID</TD>
                <TD>Log % Rough Terrain</TD>
                <TD>Est. Deaths</TD>

            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Afghanistan</TD>
                <TD bgcolor="#FFFFFF">700</TD>
                <TD bgcolor="#FFFFFF">4.2</TD>
                <TD bgcolor="#FFFFFF">54753</TD>

                
            </TR>
            <TR>
                <TD colspan ="4">Albania dropped!</TD>



            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Algeria</TD>
                <TD bgcolor="#FFFFFF">615</TD>
                <TD bgcolor="#FFFFFF">2.8</TD>
                <TD bgcolor="#FFFFFF">18488</TD>

            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Angola</TD>
                <TD bgcolor="#FFFFFF">540</TD>
                <TD bgcolor="#FFFFFF">2.4</TD>
                <TD bgcolor="#FFFFFF">23088</TD>

            </TR>
            <TR><TD colspan ="4"> ...more rows</TD></TR>


        </TABLE>
    >]

        subgraph cluster_split {
        label="";
        style=filled;
        color=white;
        node [shape=box, style=filled, fillcolor=white];
        rank=same;
        civil_wars
        rugged 
        

     }



    
    
    civil_wars -> joined
    
    rugged -> joined

   
}

```


## Full Join

Full joins will preserve all keys in both data sets. (so Albania is kept, but with an NA in the deaths column)


```{dot}

digraph G {
    rankdir=TD;

    node [shape=box];

 "rugged" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4">Terrain (LHS)</TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>Country ID</TD>
                <TD>Log % Rough Terrain</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Afghanistan</TD>
                <TD bgcolor="#FFFFFF">700</TD>
                <TD bgcolor="#FFFFFF">4.2</TD>

                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Albania</TD>
                <TD bgcolor="#FFFFFF">339</TD>
                <TD bgcolor="#FFFFFF">2.3</TD>

            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Algeria</TD>
                <TD bgcolor="#FFFFFF">615</TD>
                <TD bgcolor="#FFFFFF">2.8</TD>

            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Angola</TD>
                <TD bgcolor="#FFFFFF">540</TD>
                <TD bgcolor="#FFFFFF">2.4</TD>

            </TR>
            <TR><TD colspan ="4"> ...more rows</TD></TR>


        </TABLE>
    >]

   "civil_wars" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4">Civil Wars (RHS)</TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>Country ID</TD>
                <TD>Est. Deaths</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Afghanistan</TD>
                <TD bgcolor="#FFFFFF">700</TD>
                <TD bgcolor="#FFFFFF">54753</TD>
                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Algeria</TD>
                <TD bgcolor="#FFFFFF">615</TD>
                <TD bgcolor="#FFFFFF">18488</TD>
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Angola</TD>
                <TD bgcolor="#FFFFFF">540</TD>
                <TD bgcolor="#FFFFFF">23088</TD>
            </TR>
            <TR><TD colspan ="4"> ...more rows</TD></TR>


        </TABLE>
    >]
    
    
      
    
    
        "joined" [label=<
        <TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0">
            <TR><TD colspan ="4">Combined Data</TD></TR>

            <TR>
                <TD>Country</TD>
                <TD>Country ID</TD>
                <TD>Log % Rough Terrain</TD>
                <TD>Est. Deaths</TD>

            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Afghanistan</TD>
                <TD bgcolor="#FFFFFF">700</TD>
                <TD bgcolor="#FFFFFF">4.2</TD>
                <TD bgcolor="#FFFFFF">54753</TD>

                
            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Albania</TD>
                <TD bgcolor="#FFFFFF">339</TD>
                <TD bgcolor="#FFFFFF">2.3</TD>
                <TD bgcolor="#FFFFFF">NA</TD>


            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Algeria</TD>
                <TD bgcolor="#FFFFFF">615</TD>
                <TD bgcolor="#FFFFFF">2.8</TD>
                <TD bgcolor="#FFFFFF">18488</TD>

            </TR>
            <TR>
                <TD bgcolor="#FFFFFF">Angola</TD>
                <TD bgcolor="#FFFFFF">540</TD>
                <TD bgcolor="#FFFFFF">2.4</TD>
                <TD bgcolor="#FFFFFF">23088</TD>

            </TR>
            <TR><TD colspan ="4"> ...more rows</TD></TR>


        </TABLE>
    >]

        subgraph cluster_split {
        label="";
        style=filled;
        color=white;
        node [shape=box, style=filled, fillcolor=white];
        rank=same;

        rugged 
        civil_wars

     }



    
    rugged -> joined
    
    civil_wars -> joined

   
}


```

## Joins in dplyr

`left_join(left_table, right_table, by = join_by(left_key == right_key))`


## Back to the original problem


:::::: columns
:::: {.column width="40%"}

::: incremental

- put the terrain data first and use a `left_join`

- join on the Gleditsch Ward country ID. (country names are not very reliable!)

- We'll want to fill `NA` values in `deaths` with "0" because these states had no civil war deaths.

:::

:::


:::: {.column width="60%"}

```{r, eval=TRUE}


joined_data<-left_join(rugged, 
                       civil_wars, 
                       by=join_by(gwcode ==gwno_a))

joined_data

```


:::

:::::



## Back to the original problem



:::::: columns
:::: {.column width="40%"}

- We can use `replace_na` to replace NA values with 0s

- And we'll just create an indicator that's `TRUE` if a country had 0 civil wars


::::



:::: {.column width="60%"}

```{r, eval=TRUE}


joined_data<-left_join(rugged,
                       civil_wars, 
                       by=join_by(gwcode ==gwno_a))|>
  mutate(deaths = replace_na(deaths, 0),
         any_civil_war = deaths>0
         )

joined_data

```


::::

:::::


## Back to the original problem


:::::: columns
::: {.column width="40%"}


```{r}

lm(any_civil_war ~ rough, data = joined_data)|>
  broom::tidy()

```

:::

::: {.column width="40%"}


```{r}




ggplot(joined_data, aes(x=any_civil_war, y=rough)) + 
  geom_boxplot() +
  theme_bw() +
  ylab("natural log of % mountainous terrain") +
  xlab("Any civil wars after 1989?")




```

:::

:::::


## Other joins

-   `inner_join` will drop any rows that don't have a match in either data set

-   `full_join` will avoid dropping any rows from either dataset.

-   `anti_join` will find rows in the right hand side without a match on the left. (you can use this to find discrepancies between groups)

-   `fuzzy_join` (part of a separate package called `fuzzyjoin`) will allow imperfect matches like slight misspellings.

