---
title: "Homework 4: Data manipulation and plotting"
format:
  html:
    toc: true
    df-print: kable
    code-tools: true
    embed-resources: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

```


You'll use data from the most recent round of the Crowd Counting Consortium to answer the questions below. The code chunk below will import the data into R. You can find a codebook with additional information on the variables at [this link](https://dataverse.harvard.edu/file.xhtml?fileId=10808412&version=9.0).



```{r}
library(tidyverse)

url<-'https://dataverse.harvard.edu/api/access/datafile/12027896'
ccc<-readr::read_csv(url)


```


## Question 1


Use `filter` and `str_detect` to get a weekly count the number of events where the `claims_summary` column mentions Immigration and Customs Enforcement (ICE). Create a line plot to visualize the trend.



```{r}
# Code here
ice_events <- ccc |>
  filter(str_detect(
    str_to_lower(claims_summary),
    "\\bice\\b|immigration and customs enforcement"
  ))|>
  mutate(week = floor_date(date, unit='week'))|>
  count(week)|>
  complete(week = seq.Date(min(week), max(week), by='week'),
           fill = list(n = 0)
           )


ggplot(ice_events, aes(x=week, y=n)) + 
  geom_line() +
  theme_bw() +
  scale_x_date(breaks='4 weeks', date_labels="%b '%y") +
  labs(x = "date",
       y = "number of events",
       title = "Total weekly events"
       )


```



## Question 2

The columns labeled `source1`, `source2`...`source30` contain up to 30 sources for each recorded event. How many events use social media sites (Instagram, Facebook, X/Twitter) as at least one of their sources?


(Consult the documentation for [tidy selection](https://tidyselect.r-lib.org/reference/starts_with.html) for some ideas on how to efficiently select multiple columns)


```{r}
# Code here

sm_count<-ccc|>
  select(starts_with("source"))|>
  mutate(id = row_number())|>
  pivot_longer(cols=starts_with('source'))|>
  drop_na()|>
  mutate(
    sm_source = str_detect(str_to_lower(value), "x.com|twitter.com|facebook.com|instragram.com"))|>
  group_by(sm_source)|>
  summarize(total = length(unique(id)))

sm_count






```




## Question 3

The `organizations` variable contains information about the organizations mentioned as participating in each event. Where there are multiple organizations, these are supplied in a semi-colon separated list.

Calculate the number of organizations involved in each event. Create a histogram showing the distribution of organizations per-event in the entire data set.

```{r}
# Code here



```




## Question 4


Are certain days of the week more likely to have protest events compared to others? How large is the disparity? Get a count of the total number of events on each week day (i.e. the total number of Monday events, Tuesday events, Wednesday events ...) and create a [bar plot](https://ggplot2.tidyverse.org/reference/geom_bar.html) to compare the average number of events for each day of the week for the entire data set.

(You'll probably want to consult the help files for either the base R `weekday` function or `lubridate::wday`.)




```{r}
# Code here




```



